---
title: "Intro to Epigenomics"
author: "Wade Boohar, updated by Jeanne Revilla"
date: "11/03/24"
output:
  pdf_document: default
  html_document:
    df_print: paged
updated: 03/08/26
---

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath("/home1/aganchet/490_cluster/analysis_data")) 
```

#Package Download and Data-cleaning
```{r}
if (!require("sesameData", quietly = TRUE))
BiocManager::install("sesameData")

if (!require("sesame", quietly = TRUE))
BiocManager::install("sesame")

if (!require("limma", quietly = TRUE))
BiocManager::install("limma")
```


#Load in all necessary packages
```{r}
library(TCGAbiolinks)
library(sesame)
library(sesameData)
library(limma)
library(ggplot2)
```

#Query, Download, and Prepare the data.
```{r}
query <- GDCquery(project = "TCGA-BRCA",
                  data.category = "DNA Methylation",
                  data.type = "Methylation Beta Value",
                  platform = "Illumina Human Methylation 450",
                  data.format = "Dataframe")
GDCdownload(query)    #<-- you will need to run this!
methylation450 <- GDCprepare(query)
```
The above data requires a LOT of memory. My recommendation: GDCdownload() and GDCprepare() in separate R sessions.
R requires a lot of memory for both.

```{r}
methylation_clinical <- as.data.frame(methylation450@colData)
betas <- as.data.frame(methylation450@assays@data@listData) #takes a longer time
cpg_sites <- as.data.frame(methylation450@rowRanges@elementMetadata)

column_mask <- ifelse(colnames(methylation_clinical) %in% c('treatments', 'primary_site', 'disease_type','sites_of_involvement'), F, T)

methylation_clinical <- methylation_clinical[,column_mask]
write.csv(methylation_clinical, 'brca_methylation_clinical.csv')


site_mask <- !grepl('-', cpg_sites$gene) & !grepl(';', cpg_sites$gene) & !is.na(cpg_sites$gene) & complete.cases(betas)
betas <- betas[site_mask,]
cpg_sites <- cpg_sites[site_mask,]

write.csv(betas, 'brca_methylation_betas.csv') #takes a longer time
write.csv(cpg_sites, 'brca_cpg_sites.csv')
```

Use the code below if you are loading in your data (i.e. you downloaded the data and closed the session, but you are back to continue working)
```{r}
methylation_clinical <- read.csv('brca_methylation_clinical.csv', row.names = 1)
betas <- read.csv('brca_methylation_betas.csv', row.names=1)
cpg_sites <- read.csv('brca_cpg_sites.csv', row.names = 1)
```


#(1) Naive Differential Methylation
```{r}
#masking out NAs
na_mask <- methylation_clinical$definition %in% c("Metastatic", "Primary solid Tumor")
methylation_clinical <- methylation_clinical[na_mask,]
betas_clean <- betas[,na_mask]

#creating our variable of interest & deciding the base group
methylation_clinical$metastatic <- ifelse(methylation_clinical$definition == "Metastatic", TRUE, FALSE)
#define your comparison and base group
methylation_clinical$metastatic <- factor(methylation_clinical$metastatic, levels = c(FALSE, TRUE))
#fitting linear models using a "target value"
mval <- t(apply(betas_clean, 1, function(x) log2(x/(1-x)))) 
#mvalue is another statistic for methylation, centered at 0 and ranges from -1 to 1
#this step will take a few seconds to run

design <- model.matrix(~ metastatic, data = methylation_clinical)
fit <- lmFit(mval, design)
fit2 <- eBayes(fit)
```

```{r}
#Extracting model into dataframe
dat <- data.frame(foldchange = fit[["coefficients"]][,2], logPvalue =  -log10(p.adjust(fit2[["p.value"]][,2],method='BY')), geneName = cpg_sites$gene)
dat$threshold <- as.factor(abs(dat$foldchange) < 1)

#Visualization
cols <- c("TRUE" = "red", "FALSE" = "blue")
ggplot(data=dat, aes(x=foldchange, y = logPvalue)) +
  geom_point(alpha=.2, size=0.6) +
  scale_colour_manual(values = cols) +
  geom_vline(xintercept = 1, colour="#990000", linetype="dashed") + 
  geom_vline(xintercept = - 1, colour="#990000", linetype="dashed") +
  geom_hline(yintercept = 2, colour = "#990000", linetype="dashed") +
  theme(legend.position="none") +
  xlab("Fold Change") +
  ylab("-log10 p value") +
  theme_bw() +
  theme(legend.position = "none")

#Saving results to a .csv file
write.csv(dat, "methylation_results_metastatic.csv")
```
#The volcano plot shows that many CpG sites are underneath the threshold for statistical signifigance. This indicates that most methylation differences between metastatic and non-metstatic patients are non significant. There is however a subset of genes in metastatic patients that are downregulated, however it is small. Furthermore, since this is a naive volcano plot, factors such as cofounders, are not controlled for, and therefore we cannot definitely say these sigificant differences are due to soly metastasis. 
#(2) Direct comparison of methylation status to transcriptional activity
...INSERT DESeq2 Stuff here to generate 'results'...
```{r}
library(DESeq2)


keep <- rna_clinical$definition %in% c("Metastatic", "Primary solid Tumor")

rna_clinical <- rna_clinical[keep, ]
rna_counts <- rna_counts[, keep]


rna_clinical$metastatic <- ifelse(rna_clinical$definition == "Metastatic", "metastatic", "non_metastatic")

rna_counts <- rna_counts[rowSums(rna_counts) >= 20, ]

head(rownames(rna_clinical))
head(colnames(rna_counts))
nrow(rna_clinical)
ncol(rna_counts)
length(intersect(rownames(rna_clinical), colnames(rna_counts)))



keep <- rna_clinical$definition %in% c("Metastatic", "Primary solid Tumor")

rna_clinical <- rna_clinical[keep, ]
rna_counts <- rna_counts[, colnames(rna_counts) %in% rownames(rna_clinical)]


rna_counts <- rna_counts[, rownames(rna_clinical)]


all(colnames(rna_counts) == rownames(rna_clinical))
dds <- DESeqDataSetFromMatrix(
  countData = rna_counts,
  colData = rna_clinical,
  design = ~ metastatic
)

dds <- DESeq(dds)
results <- results(dds)
results <- as.data.frame(results)

results$gene_name <- rna_genes$gene_name[
  match(rownames(results), rna_genes$gene_id)
]




write.csv(results, "deSeq_results_metastatic.csv")
de_results <- read.csv("deSeq_results_metastatic.csv")
rna_clinical <- read.csv("brca_rna_clinical_data.csv", row.names = 1, check.names = FALSE)
rna_genes <- read.csv("brca_rna_gene_data.csv", check.names = FALSE)
rna_counts <- read.csv("brca_rna_count_data.csv", check.names = FALSE)
file.info("brca_rna_count_data.csv")
readLines("brca_rna_count_data.csv", n = 5)

dim(rna_counts)
head(rna_counts)
rm(rna_clinical, rna_counts, de_results)

dim(rna_counts_test)
head(rna_counts_test[, 1:5])
colnames(rna_counts_test)[1:5]

rownames(rna_counts_test) <- rna_counts_test[, 1]
rna_counts <- rna_counts_test[, -1]

dim(rna_counts)
head(colnames(rna_counts))
head(rownames(rna_counts))

rna_counts <- as.data.frame(lapply(rna_counts, as.numeric))
rownames(rna_counts) <- rownames(rna_counts_test)

dim(rna_counts)
summary(rna_counts[, 1])

head(rownames(rna_clinical))
head(colnames(rna_counts))
length(intersect(rownames(rna_clinical), colnames(rna_counts)))
colnames(rna_counts) <- gsub("\\.", "-", colnames(rna_counts))

length(intersect(rownames(rna_clinical), colnames(rna_counts)))
all(rownames(rna_clinical) == colnames(rna_counts))
```


```{r}
#Extract genes of interest
de_results <- read.csv("deSeq_results_metastatic.csv")
downregulated <- de_results[(de_results$log2FoldChange < 1), 'gene_name']
hypermethylated <- dat[dat$foldchange > 1.5, 'geneName'] #what cutoff should you use here?
interest_genes <- intersect(downregulated, hypermethylated)
```

```{r}
#Explore your gene of interest
GENE<-"DNAH10"

gene_counts_mask <- rna_genes$gene_name == GENE
gene_betas_mask <- cpg_sites$gene == GENE

#What are we doing here?
rna_clinical_metastatic <- rna_clinical$definition == "Metastatic"
methylation_clinical_metastatic <- methylation_clinical$definition == "Metastatic"

rna_clinical_nonmetastatic <- rna_clinical$definition == "Primary solid Tumor"
methylation_clinical_nonmetastatic <- methylation_clinical$definition == "Primary solid Tumor"
#What are we doing here?
rna_metastatic <- as.numeric(rna_counts[gene_counts_mask, rna_clinical_metastatic])
methylation_metastatic <- (betas[gene_betas_mask, methylation_clinical_metastatic])

rna_nonmetastatic <- as.numeric(rna_counts[gene_counts_mask, rna_clinical_nonmetastatic])
methylation_nonmetastatic <- betas[gene_betas_mask, methylation_clinical_nonmetastatic]

#Show transcription differences
boxplot(rna_nonmetastatic, rna_metastatic, xlab='Group', ylab='Counts', names=c('Non-metastatic', 'Metastatic'),
        col=c("blue","red"))
```

```{r}
#Show methylation differences at cpg sites in this gene 
plot(1:dim(methylation_nonmetastatic)[1], apply(methylation_nonmetastatic, 1, mean), col='blue', type='h', lwd=3, ylab='Average Beta', xlab='CpG sites')

lines(1.3:(dim(methylation_metastatic)[1]+0.3), apply(methylation_metastatic, 1, mean), col='red', type='h', lwd=3)

legend("topright",
       legend=c("Non-metastatic", "Metastatic"),
       col=c("blue", "red"),
       lwd=3,
       cex=0.7)
```
#In both  plots, the non metastatic and metastatic groups have similar mdedian methylation and transcription values at DNAH10 and avaerage beta values for cpg sites. While the metatstic patients in methylation do have more variability, there appears to be no sigificant shift in transcription due to this. Therefore gene expression is not entirely regulated by methylation differences in the metastatic group.


#3: Analyzing DNAH10 on the UCSC Genome Browser shows that CpG islands are broaldy distributed throughout the sequence and not neccessarily concnetrated on a promoter. This supports the data above as it demonstartes how for this specific gene, methaylation isnt highly associated with gene expression. Previous literature supports this idea as well where gene-body methylation was shown to have weak effects on gene expression. https://www.oncotarget.com/article/497/text